---
title: "Seurat PBMC3K Single-Cell RNA-seq Working Example"
output:
  html_document:
    theme: united
    df_print: kable
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
---

```{r setup, include=FALSE}
all_times <- list()  # store the time for each chunk
output_dir <- Sys.getenv("PBMC3K_OUTPUT_DIR", unset = "output")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(output_dir, "images"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(output_dir, "timings"), recursive = TRUE, showWarnings = FALSE)
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res <- difftime(Sys.time(), now, units = "secs")
      all_times[[options$label]] <<- res
    }
  }
}))
knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  time_it = TRUE,
  error = FALSE
)
```

# Working Example Overview

This R Markdown file is a reproducible working example of a Seurat single-cell RNA-seq workflow. It analyzes the public 10x Genomics PBMC3K dataset: 2,700 peripheral blood mononuclear cells sequenced on an Illumina NextSeq 500. The raw data can be downloaded from 10x Genomics [here](https://cf.10xgenomics.com/samples/cell/pbmc3k/pbmc3k_filtered_gene_bc_matrices.tar.gz).

Input data are expected in a 10x feature-barcode matrix directory. Set `PBMC3K_DATA_DIR` to the local data directory, or place the matrix under `data/filtered_gene_bc_matrices/hg19/`. Outputs are written to `PBMC3K_OUTPUT_DIR` if set, otherwise to `output/`.

The rendered artifact includes QC plots, PCA/UMAP visualizations, marker-gene tables, a cell-type annotation table, saved Seurat `.rds` objects, timing information, and `sessionInfo()` for reproducibility.

# Set Up the Seurat Object

The workflow reads the 10x count matrix with `Read10X()` and stores the data in a Seurat object. In Seurat v5, the count matrix is stored in `pbmc[["RNA"]]$counts`.

```{r init}
library(dplyr)
library(Seurat)
library(patchwork)

# Load the PBMC dataset
data_dir <- Sys.getenv("PBMC3K_DATA_DIR", unset = "data/filtered_gene_bc_matrices/hg19/")
pbmc.data <- Read10X(data.dir = data_dir)
# Initialize the Seurat object with the raw, non-normalized data.
pbmc <- CreateSeuratObject(counts = pbmc.data, project = "pbmc3k", min.cells = 3, min.features = 200)
pbmc
```
<details>
  <summary>**Count matrix preview**</summary>

```{r}
# Examine a few genes in the first thirty cells.
pbmc.data[c("CD3D","TCL1A","MS4A1"), 1:30]
```

The `.` values in the matrix represent 0s (no molecules detected). Since most values in an scRNA-seq matrix are 0,  Seurat uses a sparse-matrix representation whenever possible. This results in significant memory and speed savings for Drop-seq/inDrop/10x data.

```{r}
dense.size <- object.size(as.matrix(pbmc.data))
dense.size
sparse.size <- object.size(pbmc.data)
sparse.size
dense.size / sparse.size
```
</details>
\  

# Standard Preprocessing Workflow

The steps below encompass the standard pre-processing workflow for scRNA-seq data in Seurat. These represent the selection and filtration of cells based on QC metrics, data normalization and scaling, and the detection of highly variable features.

## QC and selecting cells for further analysis

The example computes common QC metrics and filters cells with documented thresholds. Frequently used QC metrics include:

* The number of unique genes detected in each cell. 
    + Low-quality cells or empty droplets will often have very few genes
    + Cell doublets or multiplets may exhibit an aberrantly high gene count
* Similarly, the total number of molecules detected within a cell (correlates strongly with unique genes)
* The percentage of reads that map to the mitochondrial genome
    + Low-quality / dying cells often exhibit extensive mitochondrial contamination
    + Mitochondrial QC metrics are calculated with `PercentageFeatureSet()`
    + The mitochondrial gene set is defined as genes starting with `MT-`

```{r mito}
# The [[ operator can add columns to object metadata. This is a great place to stash QC stats
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")
```

<details>
  <summary>**Where are QC metrics stored in Seurat?**</summary>

* The number of unique genes and total molecules are automatically calculated during `CreateSeuratObject()`
    + You can find them stored in the object meta data
```{r qc}
# Show QC metrics for the first 5 cells
head(pbmc@meta.data, 5)
```
</details>
\  

The workflow visualizes QC metrics and applies the following filtering criteria:

* Keep cells with 200 to 2,500 detected features
* Remove cells with >5% mitochondrial counts
    
```{r qc2, fig.height=7, fig.width=13}

# Visualize QC metrics as a violin plot.
VlnPlot(pbmc, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)

# FeatureScatter is typically used to visualize feature-feature relationships, but can be used for anything calculated by the object, i.e. columns in object metadata, PC scores etc.

plot1 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "percent.mt") 
plot2 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "nFeature_RNA") 
plot1 + plot2

pbmc <- subset(pbmc, subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5)
```

***

# Normalize the Data

After filtering, the workflow applies Seurat's `LogNormalize` method, which normalizes each cell by total expression, multiplies by a scale factor of 10,000, and log-transforms the result. In Seurat v5, normalized values are stored in `pbmc[["RNA"]]$data`.



```{r normalize}
pbmc <- NormalizeData(pbmc, normalization.method = "LogNormalize", scale.factor = 1e4)
```
The same normalization can be run with Seurat defaults:

```{r normalize.default, eval = FALSE}
pbmc <- NormalizeData(pbmc)
```

For datasets where global-scaling assumptions are inappropriate, `SCTransform()` is a common alternative normalization path. This working example keeps the standard `LogNormalize` path so that the full workflow is compact and easy to rerun.


# Identify Highly Variable Features

The workflow identifies genes with high cell-to-cell variation and uses them for downstream dimensionality reduction and clustering.

`FindVariableFeatures()` returns the top 2,000 variable features by default. Here, the same value is specified explicitly for reproducibility.

```{r var_features, fig.height=5, fig.width=10}
pbmc <- FindVariableFeatures(pbmc, selection.method = 'vst', nfeatures = 2000)

# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(pbmc), 10)

# Plot variable features with and without labels.
plot1 <- VariableFeaturePlot(pbmc)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot1 + plot2
```

***

# Scale the Data

The workflow applies a linear transformation ('scaling') prior to dimensional reduction. The `ScaleData()` function:

* Shifts the expression of each gene, so that the mean expression across cells is 0
* Scales the expression of each gene, so that the variance across cells is 1
    + This step gives equal weight in downstream analyses, so that highly-expressed genes do not dominate
* The results of this are stored in `pbmc[["RNA"]]$scale.data`
* By default, only variable features are scaled. 
* You can specify the `features` argument to scale additional features

```{r regress, results='hide'}
all.genes <- rownames(pbmc)
pbmc <- ScaleData(pbmc, features = all.genes)
```

<details>
  <summary>**Optional covariate regression**</summary>

`ScaleData()` can regress out unwanted sources of variation, such as cell-cycle state or mitochondrial contamination:
```{r regressvarmt, eval = FALSE}
pbmc <- ScaleData(pbmc, vars.to.regress = 'percent.mt')
```
For larger or more complex datasets, `SCTransform()` can also be used with `vars.to.regress`.
</details>

***

# Perform Linear Dimensional Reduction

The workflow performs PCA on the scaled data using the previously selected variable features.

For the first principal components, Seurat outputs a list of genes with the most positive and negative loadings, representing modules of genes that exhibit either correlation (or anti-correlation) across single-cells in the dataset.

```{r pca,results='hide'}
pbmc <- RunPCA(pbmc, features = VariableFeatures(object = pbmc))
```

PCA outputs are inspected with `VizDimReduction()`, `DimPlot()`, and `DimHeatmap()`.

```{r pca_viz, fig.height=8, fig.width=10, message=TRUE}
# Examine and visualize PCA results.
print(pbmc[['pca']], dims = 1:5, nfeatures = 5)
VizDimLoadings(pbmc, dims = 1:2, reduction = 'pca')
DimPlot(pbmc, reduction = 'pca') + NoLegend()
```

`DimHeatmap()` provides a compact check of the primary sources of heterogeneity in the dataset and helps document the PC selection used for clustering.

```{r single-heatmap, fig.height=6, fig.width=9}
DimHeatmap(pbmc, dims = 1, cells = 500, balanced = TRUE)
```

```{r multi-heatmap, fig.height=15, fig.width=9}
DimHeatmap(pbmc, dims = 1:15, cells = 500, balanced = TRUE)
```

# Choose the Number of PCs

The workflow uses the first 10 PCs for neighbor finding, clustering, and UMAP. The elbow plot below documents that choice.

The elbow plot ranks principal components by variance explained. For this dataset, the elbow around PC9-10 supports using the first 10 PCs.

```{r elbow_plot, fig.height=6, fig.width=10}
ElbowPlot(pbmc)
```

For a production analysis, PC choice should be validated by checking marker genes, cluster stability, and sensitivity to reasonable alternatives. This example fixes the value at 10 so the rendered artifact is deterministic.

* Dendritic cell and NK aficionados may recognize that genes strongly associated with PCs 12 and 13 define rare immune subsets (i.e. MZB1 is a marker for plasmacytoid DCs). However, these groups are so rare, they are difficult to distinguish from background noise for a dataset of this size without prior knowledge. 
* Repeating downstream steps with 10, 15, or 50 PCs is a useful robustness check for larger analyses.
* Using too few PCs can remove meaningful biological structure.

***

# Cluster the Cells

Seurat  applies a graph-based clustering approach, building upon initial strategies in ([Macosko *et al*](http://www.cell.com/abstract/S0092-8674(15)00549-8)). Importantly, the *distance metric* which drives the clustering analysis (based on previously identified PCs) remains the same. However, our approach to partitioning the cellular distance matrix into clusters has dramatically improved. Our approach was heavily inspired by recent manuscripts which applied graph-based clustering approaches to scRNA-seq data [[SNN-Cliq, Xu and Su, Bioinformatics, 2015]](http://bioinformatics.oxfordjournals.org/content/early/2015/02/10/bioinformatics.btv088.abstract) and CyTOF data [[PhenoGraph, Levine *et al*., Cell, 2015]](http://www.ncbi.nlm.nih.gov/pubmed/26095251). Briefly, these methods embed cells in a graph structure - for example a K-nearest neighbor (KNN) graph, with edges drawn between cells with similar feature expression patterns, and then attempt to partition this graph into highly interconnected 'quasi-cliques' or 'communities'. 

The workflow constructs a KNN graph in PCA space and refines edge weights using shared neighborhood overlap. This step is performed with `FindNeighbors()` using the first 10 PCs.

`FindClusters()` applies modularity optimization. This working example uses resolution 0.5 and stores the resulting cluster assignments in the Seurat object.


```{r cluster}
pbmc <- FindNeighbors(pbmc, dims = 1:10)
pbmc <- FindClusters(pbmc, resolution = 0.5)

# Look at cluster IDs of the first 5 cells
head(Idents(pbmc), 5)
```

***

# Run Nonlinear Dimensional Reduction

Seurat offers several non-linear dimensional reduction techniques, such as tSNE and UMAP, to visualize and explore these datasets. The goal of these algorithms is to learn underlying structure in the dataset, in order to place similar cells together in low-dimensional space. Therefore, cells that are grouped together within graph-based clusters determined above should co-localize on these dimension reduction plots. 

UMAP is used here as a visualization and QC aid. Biological conclusions should be supported by marker genes, metadata, and quantitative analyses rather than UMAP position alone.



```{r umap}
pbmc <- RunUMAP(pbmc, dims = 1:10)
```

```{r umapplot, fig.height=8, fig.width=8}
# Cluster labels can be added with `label = TRUE` or `LabelClusters()`.
DimPlot(pbmc, reduction = 'umap')
```

The intermediate object is saved so downstream steps can be rerun without recomputing preprocessing, PCA, clustering, and UMAP.

```{r saveobject}
saveRDS(pbmc, file = file.path(output_dir, "pbmc3k_intermediate.rds"))
```

***

# Find Differentially Expressed Features

Seurat can help you find markers that define clusters via differential expression (DE). By default, it identifies positive and negative markers of a single cluster (specified in `ident.1`), compared to all other cells.  `FindAllMarkers()` automates this process for all clusters, but you can also test groups of clusters vs. each other, or against all cells.

If the `presto` package is installed, Seurat v5 can use it to speed up differential expression analysis on large datasets. The marker table generated below is exported as a reproducible output.

```{r markers1}
# find all markers of cluster 2
cluster2.markers <- FindMarkers(pbmc, ident.1 = 2)
head(cluster2.markers, n = 5)
# find all markers distinguishing cluster 5 from clusters 0 and 3
cluster5.markers <- FindMarkers(pbmc, ident.1 = 5, ident.2 = c(0, 3))
head(cluster5.markers, n = 5)
# find markers for every cluster compared to all remaining cells, report only the positive ones
pbmc.markers <- FindAllMarkers(pbmc, only.pos = TRUE)
write.csv(pbmc.markers, file = file.path(output_dir, "pbmc3k_cluster_markers.csv"), row.names = FALSE)
pbmc.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1)
```

Seurat has several tests for differential expression that can be set with the `test.use` parameter. The ROC test returns the classification power for an individual marker, ranging from 0 for random to 1 for perfect.

```{r markersroc}
cluster0.markers <- FindMarkers(pbmc, ident.1 = 0, logfc.threshold = 0.25, test.use = "roc", only.pos = TRUE)
```

Marker expression is visualized with `VlnPlot()`, `FeaturePlot()`, and `DoHeatmap()`.

```{r markerplots, fig.height=5, fig.width=10}
VlnPlot(pbmc, features = c("MS4A1", "CD79A"))
# Raw counts can also be plotted.
VlnPlot(pbmc, features = c("NKG7", "PF4"), slot = 'counts', log = TRUE)
```

```{r, fig.height=8, fig.width=10}
FeaturePlot(pbmc, features = c("MS4A1", "GNLY", "CD3E", "CD14", "FCER1A", "FCGR3A", "LYZ", "PPBP", "CD8A"))
```
`DoHeatmap()` generates an expression heatmap for given cells and features. In this case, we are plotting the top 10 markers (or all markers if less than 10) for each cluster.

```{r clusterHeatmap, fig.height=8, fig.width=10}
pbmc.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 10) %>%
    ungroup() -> top10
DoHeatmap(pbmc, features = top10$gene) + NoLegend()
```

***
# Assign Cell-Type Identity to Clusters

Canonical PBMC marker genes are used to assign cell-type labels to the unsupervised clusters:

Cluster ID | Markers       | Cell Type
-----------|---------------|----------
0          | IL7R, CCR7    | Naive CD4+ T 
1          | CD14, LYZ     | CD14+ Mono
2          | IL7R, S100A4  | Memory CD4+ 
3          | MS4A1         | B 
4          | CD8A          | CD8+ T 
5          | FCGR3A, MS4A7 | FCGR3A+ Mono
6          | GNLY, NKG7    | NK 
7          | FCER1A, CST3  | DC
8          | PPBP          | Platelet


```{r labelplot, fig.height=5, fig.width=9}
new.cluster.ids <- c("Naive CD4 T", "CD14+ Mono", "Memory CD4 T", "B", "CD8 T", "FCGR3A+ Mono", "NK", "DC", "Platelet")
names(new.cluster.ids) <- levels(pbmc)
annotation_table <- data.frame(cluster_id = names(new.cluster.ids), cell_type = unname(new.cluster.ids))
write.csv(annotation_table, file = file.path(output_dir, "pbmc3k_cell_type_annotations.csv"), row.names = FALSE)
pbmc <- RenameIdents(pbmc, new.cluster.ids)
DimPlot(pbmc, reduction = 'umap', label = TRUE, pt.size = 0.5) + NoLegend()
```

```{r save.img, include=TRUE}
library(ggplot2)
plot <- DimPlot(pbmc, reduction = "umap", label = TRUE, label.size = 4.5) + xlab("UMAP 1") + ylab("UMAP 2") + 
  theme(axis.title = element_text(size = 18), legend.text = element_text(size = 18)) + 
  guides(colour = guide_legend(override.aes = list(size = 10)))
ggsave(filename = file.path(output_dir, "images", "pbmc3k_umap.jpg"), height = 7, width = 12, plot = plot, quality = 50)
```

```{r save.rds}
saveRDS(pbmc, file = file.path(output_dir, "pbmc3k_final.rds"))
```

```{r save.times, include = FALSE}
write.csv(x = t(as.data.frame(all_times)), file = file.path(output_dir, "timings", "pbmc3k_working_example_times.csv"))
```

<details>
  <summary>**Session Info**</summary>
```{r session_info}
writeLines(capture.output(sessionInfo()), con = file.path(output_dir, "sessionInfo.txt"))
sessionInfo()
```
</details>
